Supplementary materials for ``Inattentive responding can induce spurious associations between task behavior and symptom measures''

### 2023-06-21 manuscripts/supp.tex
manuscript proofs

% specify some custom commands
\renewcommand{\baselinestretch}{1.2}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand*{\Affilfont}{\normalsize}

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1,*]{Samuel Zorowitz}
\author[2]{Johanne Solis}
\author[1,3]{Yael Niv}
\author[4]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, NJ, USA}
\affil[2]{Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry, Rutgers University, NJ, USA}
\affil[3]{Department of Psychology, Princeton University, NJ, USA}
\affil[4]{School of Psychological Sciences, Monash University, Victoria, Australia}
\affil[*]{Corresponding author (zorowitz@princeton.edu)}

% specify the title
\title{Supplementary materials for ``Inattentive responding can induce spurious associations between task behavior and symptom measures''}

@@

@@

\maketitle

@@

\clearpage
\section*{Appendix A}
\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

\subsection*{Signed correlation analysis}

@@

\clearpage

\subsection*{Posterior predictive checks}
\begin{figure}[H]
    \includegraphics[width=16cm]{../figures/supp_05.png}
    \centering
    \caption{Posterior predictive checks for the risk-sensitive temporal difference learning model. (A) Observed (black) and predicted (blue) learning curves averaged across the group ($N=375$). (B) Observed versus predicted choice accuracy across participants ($N=375$). (C) Distribution of average predictive density across participants ($N=375$).}
    \label{fig:figS05}
\end{figure}

\clearpage
\subsection*{Participant demographics}
\begin{table}[!h]
    \centering
    \small
    \setlength{\tabcolsep}{15pt}
    \begin{tabular}{ rrrrr }
        \toprule
        & \multicolumn{2}{c}{MTurk} & \multicolumn{2}{c}{Prolific} \\
        \cmidrule(lr){2-3} \cmidrule(lr){4-5}
        Total & \multicolumn{2}{c}{N=186} & \multicolumn{2}{c}{N=200} \\
        \midrule
        Age & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        18-25   &      16 &   8.6 &       78 &  39.0 \\
        26-35   &      76 &  40.9 &       69 &  34.5 \\
        36-45   &      46 &  24.7 &       31 &  15.5 \\
        46-55   &      22 &  11.8 &       13 &   6.5 \\
        55+     &      26 &  14.0 &        9 &   4.5 \\
        \midrule
        Gender & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        Female             &    83 &  44.6 &    112 &  56.0 \\
        Male               &   103 &  55.4 &     85 &  42.5 \\
        Other              &     0 &   0.0 &      3 &   1.5 \\
        \midrule
        Ethnicity & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        Hispanic or Latino     &    15 &   8.1 &     10 &   5.0 \\
        Not Hispanic or Latino &   168 &  90.3 &    183 &  91.5 \\
        Rather not say         &     2 &   1.1 &      7 &   3.5 \\
        Unknown                &     1 &   0.5 &      0 &   0.0 \\
        \midrule
        Race & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        African American                 &    21 &  11.3 &      7 &   3.5 \\
        Asian                                     &     5 &   2.7 &     53 &  26.5 \\
        White                                     &   151 &  81.2 &    121 &  60.5 \\
        Multiracial                               &     6 &   3.2 &      4 &   2.0 \\
        Rather not say                            &     1 &   0.5 &     12 &   6.0 \\
        \midrule
        Use other platform & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        Yes            &      71 &  38.2 &       28 &  14.0 \\
        No             &     115 &  61.8 &      172 &  86.0 \\
        \bottomrule
    \end{tabular}
    \caption{The demographics of each sample by online labor market. On average, the samples were similar though the sample from Mechanical Turk was older (two-tailed, two-sample $t$-test: $t(384)=6.567$, $p<0.001$, $d=0.669$, 95\% CI = [5.4, 10.0]) and comprised of fewer women (two-tailed, two-sample proportions test: $z(384)=2.529$, $p=0.011$, $h=0.258$, 95\% CI = [0.030, 0.228]). Note: the demographics do not include 20 participants excluded for participating in the study twice, once per platform.}
    \label{tab:tabS01}
\end{table}

@@

\clearpage
\subsection*{Task Instructions}

@@

\subsection*{C/IE responding manifests as a distinct behavioral strategy}

@@

\break
\section*{Appendix B: Replication study}

@@

\subsection*{Methods}

\subsubsection*{Participants}

@@

\subsubsection*{Experiment}

@@

\subsubsection*{Self-report measures}

@@

\subsubsection*{Correspondence of screening measures}

@@

\subsubsection*{Correlations between behavior and symptom measures}

@@

\subsection*{Results}

\subsubsection*{Careless participants (often) appear symptomatic when the overall level of symptom endorsement is low}

@@

\subsubsection*{Low correspondence between task and self-report measures of C/IE responding}

@@

\subsubsection*{Spurious symptom-behavior correlations produced by C/IE responding}

@@

\break
\subsection*{Correspondence of screening measures}

@@

\break
\section*{Appendix C: Attention checks in healthy and psychiatric patients}

@@

52 participants were recruited to participate in a behavioral experiment. Of these, 14 participated in both studies. In total, there was data from 20 sessions involving a healthy participant and data from 45 sessions involving a psychiatric participant. Of the 17 healthy participants, 6 identified as men and 11 identified as women. Of the 35 psychiatric patients, 6 identified as men, 27 identified as women, 1 identified as non-binary, and 1 preferred not to say. The healthy participants were 24.2 years old on average (sd = 3.5), whereas the psychiatric patients were 27.4 years old on average (sd = 10.6).

\subsubsection*{Infrequency items}

To measure and compare C/IE responding between healthy and psychiatric participants, we used two sets of attention checks. Each set was composed of three infrequency items and one instructed item. Participants were assigned one or the other set. The six infrequency items were:

@@

\subsubsection*{Analysis}

Of primary interest here was whether psychiatric patients fail infrequency item attention checks at equal or at greater rates than healthy participants. To test this, we conducted Bayesian hypothesis testing using Bayes factors \cite{harms2018making}. Specifically, we defined three competing models:

@@

where $N_1$ and $N_2$ are the observed number of healthy and psychiatric participants, respectively; $p$ is the latent probability of a participant failing one or more attention checks; and $\delta$ is an offset parameter specifying the hypothesized difference in failure rates between groups. Thus, $M_0$ assumes equal rates of failures between healthy and psychiatric participants, whereas $M_1$ and $M_2$  assume greater and lower rates, respectively, for psychiatric participants compared to healthy participants. The common latent probability, $p$, was set to the observed average rate across the two groups. The offset, $\delta$, was set to 0.05. This value was selected because it signifies a difference in portions of $\Delta p = 0.1$, corresponding to a small effect for a difference in proportions test ($h = 0.2$; \cite{cohen2013statistical}).

\subsection*{Results}

In total, 16 of 65 (24.6\%) sessions involved one or more failed attention checks. Interestingly, the overall proportion of flagged sessions was similar to that observed for the original study. Subdivided by group, 6 of 20 healthy participant sessions (30\%) and 10 of 45 psychiatric participant sessions (22\%) were flagged for C/IE responding. Unsurprisingly, given the modest sample size, the difference between the two proportions was not significantly different from zero (two-tailed, two-sample proportions test: $z(63)=0.672$, $p=0.502$, $h=0.178$, 95\% CI = [-0.157, 0.312]). Across the six infrequency items, the average failure rate was 8.5\% (range: 0.0\% -- 19.7\%). In contrast, no participant failed either instructed item. This result further corroborates previous research, which has found that instructed items are poor measures of C/IE responding \cite{barends2019noncompliant-2, thomas2017validity-2, hauser2016attentive-2}.

Next, we computed the Bayes factor for each pair of candidate models. A model assuming equal rates of failure between healthy and psychiatric participants was 2.88 times more likely than the model assuming greater rates for psychiatric patients. In turn, a model assuming lower rates of failure for psychiatric patients was 1.27 times more likely than the model assuming equal rates. Finally, a model assuming lower rates of failure for psychiatric patients was 3.65 times more likely than the model assuming higher rates for psychiatric patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a model comparison as meaningful.

@@

Here we sought to examine whether actual psychiatric patients were equally or more likely to fail infrequency items than healthy controls in settings similar to those experienced by online participants (i.e., completing an experiment online, on a computer in one's home or otherwise chosen environment). Although the small sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where psychiatric patients were more likely to fail infrequency-item attention checks. Indeed, a model in which healthy controls failed attention checks at a greater rate than patients was credibly preferred to the alternative. This preliminary finding may reflect differences in motivation between patients and controls for participating in psychiatric research. Indeed, whereas healthy controls may be primarily motivated to participate for monetary purposes, patients may be motivated to participate to further scientific research that may ultimately benefit them (or others suffering from the same conditions). That is, patients may have more ``stakes in the game,'' and may therefore be more motivated to provide higher-quality responses. Regardless, further research is required to examine whether this preliminary finding holds in larger samples and other testing contexts.

>>>>

% specify the title
\title{}

@@

% Define title of Table of Contents
\addto\captionsenglish{\renewcommand{\contentsname}{Table of Contents}}

@@

% \maketitle

% Change ToC link color
{\hypersetup{linkcolor=black, linkbordercolor=black} \tableofcontents}

@@

\clearpage
\addcontentsline{toc}{section}{Supplementary Materials A - Original Study}
\section*{Supplementary Materials A}
\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}

\addcontentsline{toc}{subsection}{Supplementary Figures}
\subsection*{Signed correlation analysis}

@@

\clearpage
\subsection*{Posterior predictive checks}
\begin{figure}[H]
    \includegraphics[width=16cm]{../figures/supp_05.png}
    \centering
    \caption{Posterior predictive checks for the risk-sensitive temporal difference learning model. (A) Observed (black) and predicted (blue) learning curves averaged across the group ($N=375$). (B) Observed versus predicted choice accuracy across participants ($N=375$). (C) Distribution of average predictive density across participants ($N=375$).}
    \label{fig:figS05}
\end{figure}

\clearpage
\addcontentsline{toc}{subsection}{Supplementary Tables}
\subsection*{Participant demographics}
\begin{table}[!h]
    \centering
    \small
    \setlength{\tabcolsep}{15pt}
    \begin{tabular}{ rrrrr }
        \toprule
        & \multicolumn{2}{c}{MTurk} & \multicolumn{2}{c}{Prolific} \\
        \cmidrule(lr){2-3} \cmidrule(lr){4-5}
        Total & \multicolumn{2}{c}{N=186} & \multicolumn{2}{c}{N=200} \\
        \midrule
        Age & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        18-25   &      16 &   8.6 &       78 &  39.0 \\
        26-35   &      76 &  40.9 &       69 &  34.5 \\
        36-45   &      46 &  24.7 &       31 &  15.5 \\
        46-55   &      22 &  11.8 &       13 &   6.5 \\
        55+     &      26 &  14.0 &        9 &   4.5 \\
        \midrule
        Gender & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        Female             &    83 &  44.6 &    112 &  56.0 \\
        Male               &   103 &  55.4 &     85 &  42.5 \\
        Other              &     0 &   0.0 &      3 &   1.5 \\
        \midrule
        Ethnicity & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        Hispanic or Latino     &    15 &   8.1 &     10 &   5.0 \\
        Not Hispanic or Latino &   168 &  90.3 &    183 &  91.5 \\
        Rather not say         &     2 &   1.1 &      7 &   3.5 \\
        Unknown                &     1 &   0.5 &      0 &   0.0 \\
        \midrule
        Race & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        African American                 &    21 &  11.3 &      7 &   3.5 \\
        Asian                                     &     5 &   2.7 &     53 &  26.5 \\
        White                                     &   151 &  81.2 &    121 &  60.5 \\
        Multiracial                               &     6 &   3.2 &      4 &   2.0 \\
        Rather not say                            &     1 &   0.5 &     12 &   6.0 \\
        \midrule
        Use other platform & N & \% & N & \% \\
        \cmidrule(lr){1-5}
        Yes            &      71 &  38.2 &       28 &  14.0 \\
        No             &     115 &  61.8 &      172 &  86.0 \\
        \bottomrule
    \end{tabular}
    \caption{The demographics of each sample by online labor market. On average, the samples were similar though the sample from Mechanical Turk was older (two-tailed, two-sample $t$-test: $t(384)=6.567$, $p<0.001$, $d=0.669$, 95\% CI = [5.4, 10.0]) and comprised of fewer women (two-tailed, two-sample proportions test: $z(384)=2.529$, $p=0.011$, $h=0.258$, 95\% CI = [0.030, 0.228]). Note: the demographics do not include 20 participants excluded for participating in the study twice, once per platform.}
    \label{tab:tabS01}
\end{table}

@@

\clearpage
\addcontentsline{toc}{subsection}{Reversal learning task instructions}
\subsection*{Reversal learning task Instructions}

@@

\addcontentsline{toc}{subsection}{Model-based analyses of C/IE behavior}
\subsection*{C/IE responding manifests as a distinct behavioral strategy}

@@

\break
\addcontentsline{toc}{section}{Supplementary Materials B - Replication Study}
\section*{Supplementary Materials B: Replication study}

@@

\addcontentsline{toc}{subsection}{Methods}
\subsection*{Methods}

\addcontentsline{toc}{subsubsection}{Participants}
\subsubsection*{Participants}

@@

\addcontentsline{toc}{subsubsection}{Experiment}
\subsubsection*{Experiment}

@@

\addcontentsline{toc}{subsubsection}{Self-report measures}
\subsubsection*{Self-report measures}

@@

\addcontentsline{toc}{subsubsection}{Screening measures correspondence analysis}
\subsubsection*{Correspondence of screening measures}

@@

\addcontentsline{toc}{subsubsection}{Correlation analyses}
\subsubsection*{Correlations between behavior and symptom measures}

@@

\addcontentsline{toc}{subsection}{Results}
\subsection*{Results}

\addcontentsline{toc}{subsubsection}{Proportion and characterization of C/IE participants}
\subsubsection*{Careless participants (often) appear symptomatic when the overall level of symptom endorsement is low}

@@

\addcontentsline{toc}{subsubsection}{Screening measures correspondence analysis}
\subsubsection*{Low correspondence between task and self-report measures of C/IE responding}

@@

\addcontentsline{toc}{subsubsection}{Correlation analyses}
\subsubsection*{Spurious symptom-behavior correlations produced by C/IE responding}

@@

\break

\addcontentsline{toc}{subsection}{Supplementary Tables}
\subsection*{Correspondence of screening measures}

@@

\break
\addcontentsline{toc}{section}{Supplementary Materials C - Clinical Study}
\section*{Supplementary Materials C: Attention checks in healthy and depressed patients}

@@

52 participants were recruited to participate in a behavioral experiment. Of these, 14 participated in both studies. In total, there was data from 20 sessions involving a healthy participant and data from 45 sessions involving a patient with major depressive disorder. Of the 17 healthy participants, 6 identified as men and 11 identified as women. Of the 35 psychiatric patients, 6 identified as men, 27 identified as women, 1 identified as non-binary, and 1 preferred not to say. The healthy participants were 24.2 years old on average (sd = 3.5), whereas the psychiatric patients were 27.4 years old on average (sd = 10.6).

\addcontentsline{toc}{subsubsection}{Attention checks}
\subsubsection*{Infrequency items}

To measure and compare C/IE responding between healthy and MDD patients, we used two sets of attention checks. Each set was composed of three infrequency items and one instructed item. Participants were assigned one or the other set. The six infrequency items were:

@@

\addcontentsline{toc}{subsubsection}{Analysis}
\subsubsection*{Analysis}

Of primary interest here was whether MDD patients fail infrequency item attention checks at equal or at greater rates than healthy participants. To test this, we conducted Bayesian hypothesis testing using Bayes factors \cite{harms2018making}. Specifically, we defined three competing models:

@@

where $N_1$ and $N_2$ are the observed number of healthy and MDD participants, respectively; $p$ is the latent probability of a participant failing one or more attention checks; and $\delta$ is an offset parameter specifying the hypothesized difference in failure rates between groups. Thus, $M_0$ assumes equal rates of failures between healthy and MDD participants, whereas $M_1$ and $M_2$ assume greater and lower rates, respectively, for MDD participants compared to healthy participants. The common latent probability, $p$, was set to the observed average rate across the two groups. The offset, $\delta$, was set to 0.05. This value was selected because it signifies a difference in portions of $\Delta p = 0.1$, corresponding to a small effect for a difference in proportions test ($h = 0.2$; \cite{cohen2013statistical}).

\addcontentsline{toc}{subsection}{Results}
\subsection*{Results}

In total, 16 of 65 (24.6\%) sessions involved one or more failed attention checks. Interestingly, the overall proportion of flagged sessions was similar to that observed for the original study. Subdivided by group, 6 of 20 healthy participant sessions (30\%) and 10 of 45 MDD participant sessions (22\%) were flagged for C/IE responding. Unsurprisingly, given the modest sample size, the difference between the two proportions was not significantly different from zero (two-tailed, two-sample proportions test: $z(63)=0.672$, $p=0.502$, $h=0.178$, 95\% CI = [-0.157, 0.312]). Across the six infrequency items, the average failure rate was 8.5\% (range: 0.0\% -- 19.7\%). In contrast, no participant failed either instructed item. This result further corroborates previous research, which has found that instructed items are poor measures of C/IE responding \cite{barends2019noncompliant-2, thomas2017validity-2, hauser2016attentive-2}.

Next, we computed the Bayes factor for each pair of candidate models. A model assuming equal rates of failure between healthy and MDD participants was 2.88 times more likely than the model assuming greater rates for MDD patients. In turn, a model assuming lower rates of failure for MDD patients was 1.27 times more likely than the model assuming equal rates. Finally, a model assuming lower rates of failure for MDD patients was 3.65 times more likely than the model assuming higher rates for MDD patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a model comparison as meaningful.

@@

Here we sought to examine whether actual MDD patients were equally or more likely to fail infrequency items than healthy controls in settings similar to those experienced by online participants (i.e., completing an experiment online, on a computer in one's home or otherwise chosen environment). Although the small sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where MDD patients were more likely to fail infrequency-item attention checks. Indeed, a model in which healthy controls failed attention checks at a greater rate than patients was credibly preferred to the alternative. This preliminary finding may reflect differences in motivation between patients and controls for participating in psychiatric research. Indeed, whereas healthy controls may be primarily motivated to participate for monetary purposes, patients may be motivated to participate to further scientific research that may ultimately benefit them (or others suffering from the same conditions). That is, patients may have more ``stakes in the game,'' and may therefore be more motivated to provide higher-quality responses. Regardless, further research is required to examine whether this preliminary finding holds in larger samples and other testing contexts.
